Think deep for reef resilience

### 2025-07-21 paper/paper_draft.qmd
Various grammar and content changes to improve flow

Understanding the characteristics of reefs that will persist under projected climate
conditions requires examining reef resilience across multiple scales and processes.
Structural resilience - the capacity to maintain coral populations and avoid ecosystem
regime shifts - depends on three interconnected mechanisms. First, the ability of individual
corals to survival thermal stress. Second, the connectivity networks that enable
recolonisation after disturbances. Third, the maintenance of carbonate framework that
provides reef structure. We aim to assess how these mechanisms interact under current
projected climate conditions to identify the characteristics of structurally resilient reefs through scenario-based coral population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of
thousands of individual reefs. They are interconnected through coral and fish larval
dispersal patterns mediated by hydrodynamic, spatial, and ecological processes
[@grimaldi2022]. The GBR faces severe projected coral cover declines under climate change [@bozec]. These coral cover declines will be compounded by ocean acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks
[@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24
Summer mass bleaching event may be signalling a shift towards this reef future, with some
level of bleaching recorded on 79% of surveyed reefs [@cantin2024].

At the individual coral scale, survival during thermal stress events depends on both
environmental and biological factors. Environmental factors can either amplify or reduce
thermal stress exposure. Depth has been suggested as one crucial factor that can provide
a refuge from thermal stress [@álvarez-noriega2018; @baird2018; @smith2014]. Biological
factors including coral size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005],
and thermal tolerance through acclimation and inheritance [@humanes2022; @lachs2024] all
influence survival outcomes. Importantly, corals demonstrate adaptive capacity to develop
thermal tolerance that could delay widespread bleaching under moderate emissions scenarios
[@bairos-novak2021; @humanes2022], representing a crucial mechanism for reef persistence.

@@

Long-term reef viability requires not only coral survival and recolonization, but also
maintenance of the physical reef structure through positive carbonate budgets. Scleractinian
corals build the primary habitat framework, but reef persistence depends on the balance
between carbonate accumulation [@perry2015] and loss through bioerosion, physical damage
[@kenyon], and dissolution [@eyre2014; @eyre2018]. This balance is influenced by coral cover
and complexity [@perry2012; @desbiens], environmental processes, and ocean chemistry
[@perry2015; @eyre2018]. For Great Barrier Reef mid and outer shelf reefs, maintaining
positive carbonate budgets requires approximately 17% live coral cover [@desbiens].

Recent advances in whole-ecosystem larval connectivity modelling and coral thermal response
research enable integrated investigation of how these multiple factors interact to determine
climate resilience. By incorporating the latest coral population models
[@ribeirodealmeida2024], ecosystem-wide connectivity data, and updated climate projections
[@dixon2022], we can examine which factors most strongly influence reef persistence under
different scenarios. While connectivity has been emphasized as critical for GBR resilience
[@hock2017], and remains an important consideration, our integrated modelling reveals depth
may play a fundamental role in determining maintenance of reef coral population and positive
carbonate budgets under current projections of climate change.

@@

The spatial extent of the GBR coral reefs represented in this investigation follows the reef
definitions available via the Great Barrier Reef Marine Park Authority
\[https://geohub-gbrmpa.hub.arcgis.com/datasets/GBRMPA::great-barrier-reef-features-/explore?location=-19.221541%2C145.776418%2C4.94\].
In total this covers 3806 reef areas across the GBR marine park. Annually and spatially
averaged Degree Heating Week (DHW) data were used to represent future thermal stress at each
reef location.

Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the Great
Barrier Reef were obtained from the HighResCoralStress database [@dixon2022; @dixon].
Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology
[@liu2017; @skirving2020]. DHWs were represented by a selection of five Global Climate
Models (GCMs) from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6;
**citation**). These five GCMs were EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and
GDFL-CM4, and their inclusion were based on evaluation by Grose et al. [-@grose2023]. Grose
et al. [-@grose2023] found high correlation between observed temperature, rainfall, El Niño
Southern Oscillation and Indian Ocean Dipole patterns in the Australian terrestrial region,
and the corresponding mechanisms in these GCMs.
<!-- @Ben, the last sentence above is quite long and doesn't explain why the five GCMs
were selected. -->

Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming
caused by a doubling of carbon dioxide above its pre-industrial concentration
[@forster2021]. The five GCMs span a range of ECS values within the IPCC assessed likely
range, and GCM development families, allowing for a range of possible climate futures to be
explored, capturing the uncertainty in future responses to CO2 increases (@fig-ecs;
[@grose2023; @shepherd2018]). The following analyses were performed using each GCM
separately. SSP 2-4.5 was used as a plausible emission pathway under current emission
trajectories, with similar atmospheric CO2 levels remaining until 2050 [@hausfather2020;
IPCC -@climate2023].

@@

![Equilibrium Climate Sensitivity (ECS) values in °C for the GCMs selected (point values).
The pink - dashed box represents the high-certainty ‘very likely’ range for ECS values and
the solid red box represents low-certainty ‘likely’ range for ECS values (Forster et al.,
2021).](../outputs/ADRIA_results/HighResCoralStress/figs/ecs_plot.png){#fig-ecs
width="`{julia} ecs_width`" height="500" align="center"}

@@

CoralBlox is a coral population growth and mortality model that simulates groups of corals
growing in diameter and competing for limited space over time [@ribeirodealmeida2024].
Details on how coral demographic processes are implemented in ADRIA - CoralBlox can be found
in the supplementary material \[**link ADRIA - CoralBlox**\].

@@

In ADRIA we track the thermal tolerance of each functional group and size class as a
truncated normal distribution that is updated in each timestep following growth and
recruitment. This distribution determines that proportion of the population on a reef that
is susceptible to bleaching (i.e. has a thermal threshold below the experienced DHW value).
Starting distributions for thermal tolerance levels are informed by Bairos-Novak et al.
[-@bairos-novak2021].

Tropical cyclones and impact of wave activity to reefs were not included in this
investigation. We scale coral bleaching impacts by a coefficient determined by the median
depth of each reef location, following Baird et al. [@baird2018], reducing the proportion of
coral colonies on a reef impacted by high heat stress as depth increases. Median depths for
each reef location were extracted from satellite derived bathymetry data for the Great
Barrier Reef (GBRMPA [-@gbr10gb2020]) (see discussion for more details). Due to varying
degrees of alignment between bathymetry data and the defined outlines for GBR reefs, reefs
with less than 5% coverage of valid bathymetry data were removed from this investigation.
Furthermore, reefs that were outside the GBRMPA management areas or not associated with a
bioregion were removed from the investigation.

@@

\*\* Describe process of estimating median depth for each reef \*\*

@@

ADRIA-CoralBlox was run for 4096 **sampled** scenarios for each of the five GCMs.
A median absolute coral cover timeseries for each reef was calculated for each GCM.
For timeseries clustering reef cover timeseries were scaled by the total area of the reef
to represent timeseries as a percentage of the total reef area. The proportion of each
functional group occupying the reef (group evenness) was also investigated as a metric for
timeseries analysis, but was found to not sufficiently capture the declined condition of reefs
suffering from very low absolute coral cover.

@@

Recent analyses suggest a general threshold of \~17% live coral cover in the case of the GBR
[@desbiens]. Given in situ carbonate budgets are influenced by (1) a range of processes; (2)
the variety of approaches that could be used to measure those processes; and (3) the
uncertainty surrounding these, it would be unreasonable to assume all reefs adhere to a
single precise $\beta_{0}$ value. Concurrently, it is difficult to ascertain a suitable
$\beta_{0}$ value for each individual reef.

Given the uncertainties present, the proportion of reefs that display capacity to maintain
coral cover above a given $\beta_{0}$ threshold is examined. An indicative range of 10 to
20% is used (inclusive, with a step size of 1%). The range was informed by Perry et al.,
[-@perry2015] and Desbians et al., [-@desbiens] providing plausible indications of the lower
and upper bounds respectively. Robustness of the conclusions drawn to the uncertainty in the
precise value of $\beta_{0}$ for all reefs can then be assessed by using the indicative
range.

@@

The distribution of number of years above carbonate-cover thresholds is right-skewed, with
many reefs experiencing low numbers of years above thresholds. Similarly, reef median depth
levels and natural log total connectivity strength levels follow right-skewed distributions.
Therefore, we analyse the correlation between number of years above carbonate-cover
thresholds and these characteristics using Spearman's correlation coefficient.

@@

To identify groups of reefs that are resilient, intermediate, and non-resilient, we
clustered reefs by their temporal behaviour as indicated by their coral cover over time. The
Complexity-Invariant Distance (CID) between each timeseries was calculated to identify
timeseries that behave similarly [@batista2014]. The CID corrects the Euclidean distance
between pairs of timeseries (which accounts for magnitude differences) with the complexity
of each timeseries (obtained through a measure of their variability) to ensure reefs are
clustered based on their underlying temporal dynamics rather than simply their average coral
cover. The process creates a distance matrix of n \* n dimensions (where n is the number of
timeseries in the group) that is clustered using a k-medoids clustering algorithm [cf.
@steinmann2020].

@@

As one might expect, the number of years above $\beta_{0}$ (the assumed positive carbonate
budget threshold) substantially decreases as $\beta_{0}$ increases.

Reefs with greater median depth levels maintain coral cover above indicative carbonate
budget levels for a longer time frame compared to reefs with shallower median depths
[@fig-access-esm1-5-depth-carbonate]. Reef median depth has a moderate positive correlation
with the number of years a reef maintains coral cover above the required threshold
[@fig-access-esm1-5-depth-carbonate]. Depth - carbonate budget maintainence correlation
decreases as the required carbonate budget threshold increases. This is likely due to an
increasing number of reefs being unable to maintain coral cover above threshold levels as
the threshold increases, with values becoming increasingly right-skewed. The natural log of
total connectivity strength was found to have no correlation with the number of years a reef
maintains cover above the required thresholds [@sfig-access-esm1-5-conn-carbonate]. Log
connectivity strength - carbonate budget maintainence correlation slightly decreases as the
required carbonate budget threshold increases.

@@

Each GCM shows similar relationships between reefs in clusters of ‘low’, ‘medium’ and ‘high’
coral cover and contextual analysis factors, thus further comparisons of these clusters for
each analysis factor are only shown for ACCESS-ESM1-5 (intermediate ECS scenario). Results
for the additional GCMs are displayed in the supplementary material. However, there are
substantial differences in the longevity of reef resilience and maintenance of coral cover
post 2030 between the GCMs, thus coral cover timeseries are displayed for each GCM
(displayed clustered at management-area scale to preserve coarse spatial differences in
coral trajectories) (@fig-gcm-cover-trajectories).

NorESM2-MM has the lowest Equilibrium Climate Sensitivity (2.5°C) and reefs exhibit
continuing recovery and decline dynamics up to 2075 under this possible climate future.
While recovery phases are predominantly seen in the ‘high’ cover cluster, the ‘medium’
cluster, and to a lesser degree the ‘low’ cluster, also exhibit periods of recovery and
increasing proportions of coral cover up to 2045. Under NorESM2-MM GCM the Far Northern
management area does not see coral cover maintenance as high as the three remaining
management regions. Under the second lowest ECS projection (GFDL-CM4, 2.9°C) all clusters in
each management area exhibit largely declining future trajectories with periods of recovery
from 2030 to 2040. ACCESS-ESM1-5 has an intermediate ECS value of 3.87°C which results in
intermediate levels of coral cover maintenance (between those of lower and higher ECS GCMs).
In Townsville/Whitsunday and Far Northern management areas reefs in the high cluster are
able to maintain coral cover levels around 20% to 2050 under ACCESS-ESM1-5. Under ACCESS-CM2
(4.72°C) and EC-Earth3-Veg (4.31°C) GCMs cover trajectories are largely declining in all
management areas.

Under each GCM, in all management areas, low, medium and high cluster reefs have low
differences between DHW trajectories (@sfig-gcm-dhw-trajectories). The highest peak DHW
levels occur under GFDL-CM4 between 2045 and 2050, while greater increases in overall DHW
timeseries occur under ACCESS-CM2.

@@

Across the five GCMs, 50% (1812) of reefs, at management region scale, remain in the same
‘low, medium, high’ cluster assignment. 1757 Reefs (49%) that changed clusters by one level
across the GCMs (e.g. a reef is assigned ‘low’ and ‘medium’ in different GCMs), while 46
(1%) reefs changed clusters by two levels across the GCMs (e.g. a reef is assigned ‘low’ and
‘high’ in different GCMs).

@@

Within bioregions timeseries clustering analysis was able to identify three clusters of
area-relative cover timeseries. Across many bioregions timeseries clustering identified a
‘low’ cluster containing reefs that suffer larger declines and generally have lower levels
of coral cover than the other clusters, a ‘medium’ cluster with intermediate levels of coral
cover, and a ‘high’ cluster with greater levels of cover recovery, with persistence to 2050
in some bioregions (@sfig-access-esm1-5-bioregion-timeseries;
@sfig-access-cm2-bioregion-timeseries; @sfig-ec-earth3-veg-bioregion-timeseries;
@sfig-gfdl-cm4-bioregion-timeseries; @sfig-noresm2-mm-bioregion-timeseries).

@@

Conversely, there are no consistent patterns across bioregions in reef area-relative cover
and mean timeseries DHW levels, with clusters having similar mean DHW levels in many
bioregions (@sfig-access-esm1-5-bioregion-dhwmean; @sfig-access-cm2-bioregion-dhwmean;
@sfig-ec-earth3-veg-bioregion-dhwmean; @sfig-gfdl-cm4-bioregion-dhwmean;
@sfig-noresm2-mm-bioregion-dhwmean). <!--
In some bioregions (4 for ACCESS-ESM1-5) reefs in the ‘low’ cluster have substantially lower mean thermal tolerance levels compared to reefs in other clusters. However, many bioregions do not show strong differences in thermal tolerance levels (@sfig-access-esm1-5-bioregion-dhwtol; @sfig-access-cm2-bioregion-dhwtol; @sfig-ec-earth3-veg-bioregion-dhwtol; @sfig-gfdl-cm4-bioregion-dhwtol; @sfig-noresm2-mm-bioregion-dhwtol). -->

@@

Under all GCMs most reefs are projected to suffer substantial declines in coral cover from
2025 levels. NorESM2-MM and ACCESS-ESM1-5 GCMs convey the most coral cover maintanance into
the future. Differences between low, medium and high reef clusters in cover timeseries are
most notable under these two GCMs. Maintainance of coral cover into the future is not
clearly linked to the Equilibrium Climate Sensitivity of each GCM used as ACCESS-ESM1-5 has
a higher ECS value than GFDL-CM4, with reefs experiencing large continuous declines under
GFDL-CM4.

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced,
and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude
DHWs based on SST, the proportion of the reef’s coral population impacted by heat stress
events depends on the depth of the reef. Analysis from Baird et al. [-@baird2018] is used to
calculate the degree of protection provided to a reef based on the median depth. Our results
suggest that within shallow reef ecosystems (depth \< 30m), depth contributes greatly to
reef resilience and the ability of the reef to maintain positive carbonate budgets under
future conditions. While similar findings have been demonstrated in the eastern Pacific
[@smith2014], depth refugia analyses have often focussed on the potential of mesophotic
reefs (\> 30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In
contrast, Venegas et al. [-@venegas2019] found no association between depth and the
alleviation of heat stress across a range of shallow reefs in the Pacific. Reefs at shallow
depths experience no, or few, years with coral cover exceeding levels likely required to
maintain positive carbonate budgets. Movement of reefs towards negative cabonate budgets
under increasing future thermal stress over time may contribute to phase shifts in reef
dominant $CaCO_3$ production from coral driven to coralline encrusting algae driven
production, or to bioerosion dominated systems with rapid erosion of physical reef
frameworks [@dudgeon2010; @perry2008].
<!-- The depth refugia found on a reef in turn influences the degree of correlation between coral cover timeseries and DHW levels. Reefs in low and medium clusters have lower levels of cover – DHW correlation than those in high clusters, due to their initial sharp declines in coral cover combined with low depth – refugia in the face of persistently high thermal stress these reefs do not recover post 2030, leaving little ability to decline further in periods of high thermal stress and recover in periods of low thermal stress. -->

Overall, these findings support the hypothesis that reefs at greater depths will experience
refugia benefits and more prolonged resilience under climate change compared to shallower
reefs. Additionally, [-@liu2024] suggest that rubble instability is inversely correlated
with reef depth, possibly limiting the larval recruitment capacity of shallow reefs. A
caveat to these findings is the quality of the bathymetry data used to calculate depth
refugia benefits in ADRIA. This data (GBRMPA [-@gbr10gb2020]) is derived from satellite
imagery and is limited in its alignment with the established GBRMPA reef outlines used in
the investigation to characterise reef locations. As a result, 842 reefs (22% of the 3806)
were removed from the analyses before depth comparisons due to having less than 5% coverage
of depth data. Additionally, deriving estimates of depth from satellite imagery is prone to
high degrees of uncertainty. Data adopted for the given assessment are said to be accurate
up to 15m from Mean Sea Level [@gbr10gb2020]. Furthermore, an estimated median depth is
determined for each reef, using the established GBRMPA reef outlines to provide an
indication of “true” reef extent. Therefore, the estimates of reef depths are taken as a
useful indication rather than an accurate and precise report of reef depths across the GBR.

>>>>

Understanding the characteristics of reefs that will persist under projected climate
conditions requires examining reef resilience across multiple scales and processes.
Structural resilience - the capacity to maintain coral populations and avoid ecosystem
regime shifts - depends on three interconnected mechanisms. First, the ability of individual
corals to survive thermal stress. Second, the connectivity networks that enable
recolonisation after disturbances. Third, the maintenance of the carbonate framework that
provides reef structure. We aim to assess how these mechanisms interact under projected
climate conditions to identify the characteristics of structurally resilient reefs through
scenario-based coral population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of
thousands of individual reefs. They are interconnected through coral and fish larval
dispersal patterns mediated by hydrodynamic, spatial, and ecological processes
[@grimaldi2022]. The GBR faces severe projected coral cover declines under climate change
[@bozec2022]. These coral cover declines will be compounded by ocean acidification,
intensifying tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012;
@fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24 Summer mass
bleaching event may be signalling a shift towards this reef future, with some level of
bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs) [@cantin2024].

At the individual coral scale, survival during thermal stress events depends on both
environmental and biological factors. Environmental factors can either amplify or reduce
thermal stress exposure. Depth has been suggested as one crucial factor that can provide a
refuge from thermal stress [@álvarez-noriega2018; @baird2018; @smith2014]. Biological
factors including coral size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005],
and thermal tolerance through acclimation and inheritance [@humanes2022; @lachs2024] all
influence survival outcomes. Importantly, corals demonstrate adaptive capacity to develop
thermal tolerance that could delay widespread bleaching under moderate emission scenarios
[@bairos-novak2021; @humanes2022], representing a crucial mechanism for reef persistence.

@@

Long-term reef viability requires not only coral survival and recolonization, but also
maintenance of the physical reef structure through positive carbonate budgets. Scleractinian
corals build the primary habitat framework, but reef persistence depends on the balance
between carbonate accumulation [@perry2015] and loss through bioerosion, physical damage
[@kenyon2023], and dissolution [@eyre2014; @eyre2018]. This balance is influenced by coral
cover and community complexity [@perry2012; @desbiens2021], environmental processes, and
ocean chemistry [@perry2015; @eyre2018]. For Great Barrier Reef mid and outer shelf reefs,
maintaining positive carbonate budgets requires approximately 17% live coral cover
[@desbiens2021].

Recent advances in whole-ecosystem larval connectivity modelling and coral thermal response
research enable integrated investigation of how these multiple factors interact to determine
climate resilience. By incorporating the latest coral population models
[@ribeirodealmeida2024], ecosystem-wide connectivity data, and updated climate projections
[@dixon2022], we can examine which factors most strongly influence reef persistence under
different scenarios. While connectivity has been emphasized as critical for GBR resilience
[@hock2017], and remains an important consideration, our integrated modelling reveals depth
may play a fundamental role in determining maintenance of reef coral populations and
positive carbonate budgets under current projections of climate change.

@@

The spatial extent of the GBR reefs represented in this investigation follows the reef
definitions available via the Great Barrier Reef Marine Park Authority [@greatba2022]. In
total this covers 3806 reef areas across the GBR marine park. Reefs that are outside the
GBRMPA management areas or are not associated with a GBRMPA reefal bioregion were removed
from the investigation, leaving 3531 reefs. Annually and spatially averaged Degree Heating
Week (DHW) data were used to represent future thermal stress at each reef location.

Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the GBR were
obtained from the HighResCoralStress database [@dixon2022; @dixon2023]. Maximum monthly mean
DHW levels were calculated from the SST data following NOAA methodology [@liu2017;
@skirving2020]. DHWs were represented by a selection of five Global Climate Models (GCMs)
from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6; **citation**).
These five GCMs were EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4, and
their inclusion were based on evaluation by Grose et al. [-@grose2023]. Grose et al.
[-@grose2023] found these GCMs to rank highly, among 50 GCMs, in recreation of historical atmospheric temperature and precipitation conditions in the Australasian region. While this assessment does not consider oceanic data, Grose et al. [@-grose2023] provide a region-specific selection mechanism which we use as an indicator of general performance of GCMs for the region.

Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming
caused by a doubling of carbon dioxide above its pre-industrial concentration
[@forster2021]. The five GCMs span a range of ECS values within the IPCC assessed likely
range, and GCM development families, allowing for a range of possible climate futures to be
explored, capturing the uncertainty in future responses to CO2 increases (@sfig-ecs;
[@grose2023; @shepherd2018]). The following analyses were performed using each GCM
separately. SSP 2-4.5 was used as a plausible emission pathway under current emission
trajectories, with similar atmospheric CO2 levels remaining until 2050 [@hausfather2020;
IPCC -@climate2023].

@@

@@

ADRIA.jl is a decision support platform that combines scenario analysis with biotic and
abiotic processes such as bleaching responses, thermal tolerance and flows of larvae between
reefs [@iwanaga2023]. CoralBlox.jl is a coral population growth and mortality model that
simulates groups of corals growing in diameter and competing for limited space over time
[@ribeirodealmeida2024]. We combine these tools to model the populations of corals across
the GBR. Details on how coral demographic processes are implemented in ADRIA - CoralBlox can
be found in the supplementary material.

@@

In ADRIA we track the thermal tolerance of each functional group and size class as a
truncated normal distribution that is updated in each timestep following growth and
recruitment. This distribution determines the proportion of the population on a reef that is
susceptible to bleaching (i.e. has a thermal threshold below the experienced DHW value).
Starting distributions for thermal tolerance levels are informed by Bairos-Novak et al.
[-@bairos-novak2021].

Tropical cyclones and impact of wave activity to reefs were not included in this
investigation. We scale coral bleaching impacts by a coefficient determined by the median
depth of each reef location, following Baird et al. [@baird2018], reducing the proportion of
coral colonies on a reef impacted by high heat stress as depth increases.

@@

Median depths for each reef location were extracted from satellite derived bathymetry data
for the Great Barrier Reef (GBRMPA [-@gbr10gb2020]) (see discussion for more details). Due
to varying degrees of alignment between bathymetry data and the defined outlines for GBR
reefs, reefs with less than 5% coverage of valid bathymetry data were removed from analyses
involving depth as a resilience factor.

@@

ADRIA-CoralBlox was run for 4096 **sampled** scenarios for each of the five GCMs (20,480
total model runs). A median absolute coral cover timeseries for each reef was calculated for
each GCM. For timeseries clustering reef cover timeseries were scaled by the total area of
the reef to represent cover as a percentage of the total reef area. The proportion of each
functional group occupying the reef (group evenness) was also investigated as a metric for
timeseries analysis, but was found to not sufficiently capture the declined condition of
reefs suffering from very low absolute coral cover.

@@

Recent analyses suggest a general threshold of \~17% live coral cover in the case of the GBR
[@desbiens2021]. Given in situ carbonate budgets are influenced by (1) a range of processes;
(2) the variety of approaches that could be used to measure those processes; and (3) the
uncertainty surrounding these, it would be unreasonable to assume all reefs adhere to a
single precise $\beta_{0}$ value. Concurrently, it is difficult to ascertain a suitable
$\beta_{0}$ value for each individual reef.

Given the uncertainties present, the proportion of reefs that display capacity to maintain
coral cover above a given $\beta_{0}$ threshold is examined. An indicative range of 10 to
20% is used (inclusive, with a step size of 1%). The range was informed by Perry et al.,
[-@perry2015] and Desbians et al., [-@desbiens2021] providing plausible indications of the
lower and upper bounds respectively. Robustness of the conclusions drawn to the uncertainty
in the precise value of $\beta_{0}$ for all reefs can then be assessed by using the
indicative range.

@@

@@

To identify groups of reefs that are resilient, intermediate, and non-resilient, we
clustered reefs into three clusters based on their temporal behaviour as indicated by coral
cover over time. The Complexity-Invariant Distance (CID) between each timeseries was
calculated to identify timeseries that behave similarly [@batista2014]. The CID corrects the
Euclidean distance between pairs of timeseries (which accounts for magnitude differences)
with the complexity of each timeseries (obtained through a measure of their variability) to
ensure reefs are clustered based on their underlying temporal dynamics rather than simply
their average coral cover. The process creates a distance matrix of n \* n dimensions (where
n is the number of timeseries in the group) that is clustered using a k-medoids clustering
algorithm [cf. @steinmann2020].

@@

As one might expect, the number of years experience cover above $\beta_{0}$ (the assumed
positive carbonate budget threshold) substantially decreases as $\beta_{0}$ increases. The
distribution of number of years above carbonate-cover thresholds is right-skewed, with many
reefs experiencing low numbers of years above thresholds.

Reefs with greater median depth levels maintain coral cover above indicative carbonate
budget levels for a longer time frame compared to reefs with shallower median depths
[@fig-access-esm1-5-depth-carbonate]. Reef median depth has a moderate positive correlation
with the number of years a reef maintains coral cover above the required threshold
[@fig-access-esm1-5-depth-carbonate]. Depth - carbonate budget maintainence correlation
decreases as $\beta_{0}$ increases. This is likely due to an increasing number of reefs
being unable to maintain coral cover above threshold levels as the threshold increases, with
values becoming increasingly right-skewed. The natural log of total connectivity strength
was found to have no correlation with the number of years a reef maintains cover above the
required thresholds [@sfig-access-esm1-5-conn-carbonate]. Log connectivity strength -
carbonate budget maintainence correlation slightly decreases as the required carbonate
budget threshold increases.

@@

Each GCM shows similar relationships between reef clusters and contextual analysis factors,
thus further comparisons of these clusters for each analysis factor are only shown for
ACCESS-ESM1-5 (intermediate ECS scenario). Results for the additional GCMs are displayed in
the supplementary material. However, there are substantial differences in the longevity of
reef resilience and maintenance of coral cover post 2030 between the GCMs, thus coral cover
timeseries are displayed for each GCM (displayed clustered at management-area scale to
preserve coarse spatial differences in coral trajectories) (@fig-gcm-cover-trajectories).

NorESM2-MM has the lowest Equilibrium Climate Sensitivity (2.5°C) and reefs exhibit
continuing recovery and decline dynamics up to 2075 under this possible climate future.
While recovery phases are predominantly seen in the ‘high’ cover cluster, the ‘medium’
cluster, and to a lesser degree the ‘low’ cluster, also exhibit periods of recovery and
increasing proportions of coral cover up to 2045. Under NorESM2-MM the Far Northern
management area does not see coral cover maintenance as high as the three remaining
management regions. Under the second lowest ECS projection (GFDL-CM4, 2.9°C) all clusters in
each management area exhibit largely declining future trajectories with periods of recovery
from 2030 to 2040. ACCESS-ESM1-5 has an intermediate ECS value of 3.87°C which results in
intermediate levels of coral cover maintenance (between those of lower and higher ECS GCMs).
In Townsville/Whitsunday and Far Northern management areas reefs in the high cluster are
able to maintain coral cover levels around 20% to 2050 under ACCESS-ESM1-5. Under ACCESS-CM2
(4.72°C) and EC-Earth3-Veg (4.31°C) GCMs cover trajectories are largely declining in all
management areas.

Under each GCM, in all management areas, low, medium and high cluster reefs have low
differences between DHW trajectories (@fig-gcm-dhw-trajectories). The highest peak DHW
levels occur under GFDL-CM4 between 2045 and 2050, while greater increases in overall DHW
timeseries occur under ACCESS-CM2.

@@

![Reef Degree Heating Week timeseries clustered within each management area, for each GCM
analysed. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover
between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with
intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’
clusters (blue) are clusters with the highest median area-relative cover between 2030 and
2040 within each bioregion. Solid lines represent the cluster median timeseries and bands
represent the 2.5 to 97.5 percentile confidence intervals. Each management area is
represented five times (for each GCM), with each GCM displayed
horizontally.](../outputs/ADRIA_results/HighResCoralStress/figs/GCM_dhw_timeseries_plot.png){#fig-gcm-dhw-trajectories
width="`{julia} timeseries_width`"}

```{=html}
<!-- Across the five GCMs, 50% (1812) of reefs, at management region scale, remain in the same
‘low, medium, high’ cluster assignment. 1757 Reefs (49%) that changed clusters by one level
across the GCMs (e.g. a reef is assigned ‘low’ and ‘medium’ in different GCMs), while 46
(1%) reefs changed clusters by two levels across the GCMs (e.g. a reef is assigned ‘low’ and
‘high’ in different GCMs). -->
```

@@

Across many bioregions timeseries clustering identified a ‘low’ cluster containing reefs
that suffer larger declines and generally have lower levels of coral cover than the other
clusters, a ‘medium’ cluster with intermediate levels of coral cover, and a ‘high’ cluster
with greater levels of cover recovery, with persistence to 2050 in some bioregions
(@sfig-access-esm1-5-bioregion-timeseries; @sfig-access-cm2-bioregion-timeseries;
@sfig-ec-earth3-veg-bioregion-timeseries; @sfig-gfdl-cm4-bioregion-timeseries;
@sfig-noresm2-mm-bioregion-timeseries).

@@

Conversely, there are no consistent patterns across bioregions in reef area-relative cover
and mean timeseries DHW levels, with clusters having similar mean DHW levels in many
bioregions (@sfig-access-esm1-5-bioregion-dhwmean; @sfig-access-cm2-bioregion-dhwmean;
@sfig-ec-earth3-veg-bioregion-dhwmean; @sfig-gfdl-cm4-bioregion-dhwmean;
@sfig-noresm2-mm-bioregion-dhwmean).

<!-- In some bioregions (4 for ACCESS-ESM1-5) reefs in the ‘low’ cluster have substantially 
lower mean thermal tolerance levels compared to reefs in other clusters. However, many 
bioregions do not show strong differences in thermal tolerance levels 
(@sfig-access-esm1-5-bioregion-dhwtol; @sfig-access-cm2-bioregion-dhwtol; 
@sfig-ec-earth3-veg-bioregion-dhwtol; @sfig-gfdl-cm4-bioregion-dhwtol; 
@sfig-noresm2-mm-bioregion-dhwtol). -->

@@

Under all GCMs most reefs are projected to suffer substantial declines in coral cover from
2025 levels. NorESM2-MM and ACCESS-ESM1-5 GCMs convey the most coral cover maintanance into
the future. Differences between low, medium and high reef clusters in cover timeseries are
most notable under these two GCMs. Maintainance of coral cover into the future is not
clearly linked to the Equilibrium Climate Sensitivity of each GCM used as ACCESS-ESM1-5 has
a higher ECS value than GFDL-CM4, with reefs experiencing large continuous declines under
GFDL-CM4. These large declines are likely due to the high peak thermal stress levels above
10°C-weeks experienced in the first ten years of the timeseries under GFDL-CM4.

The large differences in cover timeseries seen across the GCMs suggest that it is important
to consider this uncertainty in future conditions when assessing possible biological
responses to change. This is supported by recent literature suggesting an effective approach
to capturing climate change uncertainty is to model climate change responses using a range
of climate models separately [@shepherd2018]. Despite large differences in coral cover
across GCMs, the patterns identified between low, medium and high clusters of reefs suggest
that the importance of depth refugia is consistent across scenarios.

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced,
and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude
DHWs based on SST, the proportion of the reefs' coral population impacted by heat stress
events depends on reef depth. Our results suggest that within shallow-water reef ecosystems
(non-mesophotic, depth \< 30m), depth contributes greatly to reef resilience and the ability
of the reef to maintain positive carbonate budgets under future conditions. While similar
findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia analyses
have often focussed on the potential of mesophotic reefs (\> 30m deep) as thermal refugia
[@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019]
found no association between depth and the alleviation of heat stress across a range of
shallow reefs in the Pacific. Reefs at shallow depths experience no, or few, years with
coral cover exceeding levels likely required to maintain positive carbonate budgets.
Movement of shallow reefs towards negative cabonate budgets under increasing future thermal
stress over time may contribute to phase shifts in reef dominant CaCO~3~ production.
Shifting from coral driven to coralline encrusting algae driven production, or to bioerosion
dominated systems with rapid erosion of physical reef frameworks [@dudgeon2010; @perry2008].
While reefs at increased depths may be able to maintain coral driven CaCO~3~ production and
accumulation into the future for longer.
<!-- The depth refugia found on a reef in turn influences the degree of correlation between coral cover timeseries and DHW levels. Reefs in low and medium clusters have lower levels of cover – DHW correlation than those in high clusters, due to their initial sharp declines in coral cover combined with low depth – refugia in the face of persistently high thermal stress these reefs do not recover post 2030, leaving little ability to decline further in periods of high thermal stress and recover in periods of low thermal stress. -->
